---
title: "MT2 CheatSheet"
output: 
  html_document: 
    keep_md: yes
date: "2024-02-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
library("naniar")
library("RColorBrewer")
library("paletteer")
```
#cancel use of scientific notation
```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```

#12

#geom_bar no y axis, use FILL, ex:
```{r}
p <- homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(na.rm=T, position="dodge")
```

#Have a look at the `linedraw` theme; also how to angle the x axis and how to put titles in the middle
```{r}
p + theme_linedraw()+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

#how to put title in the middle
theme(plot.title=element_text(size=rel(1.5), hjust=.5))

#how to make thresholds 
```{r}
homerange <- homerange %>% 
             mutate(mass_category = case_when(log10.mass <= 1.75 ~ "small",
                                              log10.mass > 1.75 & log10.mass <= 2.75 ~ "medium",
                                              log10.mass > 2.75 ~ "large"))
```

3. Isolate the small `range_category` and plot the range of `log10.mass` by taxonomic class.
```{r}
homerange %>% 
  filter(range_category=="small") %>% 
  ggplot(aes(x=class, y= log10.mass, fill=class))+
  geom_boxplot()
```

1. Make a histogram of `log10.hra`. Make sure to add a title.
```{r}
homerange %>% 
  ggplot(aes(x = log10.hra)) +
  geom_histogram(color = "black", fill = "gray70")+
  labs(title = "Distribution of Home Range")
```

2. Now plot the same variable using `geom_density()`.
```{r}
homerange %>% 
  ggplot(aes(x = log10.hra)) +
  geom_density(fill="deepskyblue4", alpha  =0.4, color = "black")+
  labs(title = "Distribution of Body Mass")
```

3. Combine them both!
```{r}
homerange %>% 
  ggplot(aes(x=log10.hra)) +
  geom_histogram(aes(y = after_stat(density)), fill = "deepskyblue4", alpha = 0.4, color = "black")+
  geom_density(color = "orange")+
  labs(title = "Distribution of Homernge")
```


#11
```{r}
life_history %>% 
  count(order, sort=T) %>% 
  ggplot(aes(x=order, y=n))+
  geom_col()+
  coord_flip()
```

#Scatter plots allow for comparisons of two continuous variables. Make a scatterplot below that compares gestation time and weaning mass.
```{r}
life_history %>% 
  ggplot(aes(x=gestation, y=wean_mass))+
  geom_jitter(na.rm=T)+
  scale_y_log10()
```

## Boxplots
Box plots help us visualize a range of values. So, on the x-axis we typically have something categorical and the y-axis is the range. Let's make a box plot that compares mass across orders.
```{r}
life_history %>% 
  ggplot(aes(x=order, y=mass))+
  geom_boxplot(na.rm=T)+
  coord_flip()
```

```{r}
life_history %>% 
  filter(order!="Cetacea") %>% 
  ggplot(aes(x=order, y=mass))+
  geom_boxplot(na.rm=T)+
  coord_flip()+
  scale_y_log10()
```

Make a plot that compares age and height of elephants.
```{r}
elephants %>% 
  ggplot(aes(x=age, y=height)) + 
  geom_point() + 
  geom_smooth(method=lm, se=F)
```

#thermoregulation
We can color the points by a categorical variable.
```{r}
p+geom_point(aes(color=thermoregulation), size=1.5)
```

We can also scale all bars to a percentage.
```{r}
homerange %>% 
  ggplot(aes(x = taxon, fill = trophic.guild))+
  geom_bar(position = position_fill())+ 
  scale_y_continuous(labels = scales::percent)+
  coord_flip()
```

#9

Rules:  
+ `pivot_longer`(cols, names_to, values_to)
+ `cols` - Columns to pivot to longer format
+ `names_to` - Name of the new column; it will contain the column names of gathered columns as values
+ `values_to` - Name of the new column; it will contain the data stored in the values of gathered columns

To fix this problem, we need to reshape the table to long format while keeping track of column names and values. We do this using `pivot_longer()`. Notice that the dimensions of the data frame change.  
```{r}
heartrate %>% 
  pivot_longer(-patient, #patient will not move
               names_to = "drug", #make a new column called "drug"
               values_to="heartrate" #values moved to a new column called "heartrate"
               )
```

## Example 2: some column names are data
Some (but not all) of the column names are data. We also have NA's.
```{r}
billboard <- read_csv("data/billboard.csv")
billboard
```

Solution 1: specify a range of columns that you want to pivot.
```{r}
billboard2 <- 
  billboard %>% 
  pivot_longer(wk1:wk76, # a range of columns
               names_to = "week",
               values_to = "rank", 
               values_drop_na = TRUE #this will drop the NA's
               )
billboard2
```

Solution 2: OR, specify columns that you want to stay fixed.
```{r}
billboard3 <- 
  billboard %>% 
  pivot_longer(-c(artist, track, date.entered), #specific columns to stay fixed
               names_to = "week",
               values_to = "rank",
               values_drop_na = TRUE
               )
billboard3
```

Solution 3: identify columns by a prefix, remove the prefix and all NA's.
```{r}
billboard %>% 
   pivot_longer(
   cols = starts_with("wk"), #columns that start with "wk"
   names_to = "week",
   names_prefix = "wk",
   values_to = "rank",
   values_drop_na = TRUE)
```

We need to start by separating the patient names from their sexes. `separate()` needs to know which column you want to split, the names of the new columns, and what to look for in terms of breaks in the data.
```{r}
heartrate2 %>% 
  separate(patient, into= c("patient", "sex"), sep = "_")
```

## Practice
1. Re-examine `heartrate2`. Use `separate()` for the sexes, `pivot_longer()` to tidy, and `arrange()` to organize by patient and drug. Store this as a new object `heartrate3`.  
```{r}
heartrate3 <- heartrate2 %>% 
  separate(patient, into=c("patient", "sex"), sep="_") %>% 
  pivot_longer(-c(patient, sex),
               names_to = "drug",
               values_to = "heartrate")
heartrate3
```

2. `unite()` is the opposite of separate(). Its syntax is straightforward. You only need to give a new column name and then list the columns to combine with a separation character.  Give it a try below by recombining patient and sex from `heartrate3`.  
```{r}
heartrate3 %>% 
  unite(patient_sex, "patient", "sex", sep=" ")
```

When using `pivot_wider()` we use `names_from` to identify the variables (new column names) and `values_from` to identify the values associated with the new columns.
```{r}
tb_data %>% 
  pivot_wider(names_from = "key", #the observations under key will become new columns
              values_from = "value") #the values under value will be moved to the new columns
```





