---
title: "`summarize()` and `group_by()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Use a combination of `select()`, `filter()`, and `mutate()` to transform data frames.  
2. Use the `skimr` package to produce summaries of data.  
3. Produce clean summaries of data using `summarize()`.  
4. Use `group_by()` in combination with `summarize()` to produce grouped summaries of data.  

## Load the tidyverse and janitor
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Install `skimr`
```{r}
install.packages("skimr")
library("skimr")
```

## Load the data
For this lab, we will use the built-in data on mammal sleep patterns. From: _V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007_.
```{r}
?msleep
names(msleep)
msleep
```

We will also use [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/articles/intro.html) data in the second part of lab.
```{r}
library("palmerpenguins")
```

## dplyr Practice
Let's do a bit more practice to make sure that we understand `select()`, `filter()`, and `mutate()`. Start by building a new data frame `msleep24` from the `msleep` data that: contains the `name` and `vore` variables along with a new column called `sleep_total_24` which is the amount of time a species sleeps expressed as a proportion of a 24-hour day. Restrict the `sleep_total_24` values to less than or equal to 0.3. Arrange the output in descending order.  
```{r}
msleep24 <- msleep %>% 
  select(name,vore,sleep_total) %>% 
  mutate(sleep_total_24= sleep_total/24) %>% 
  filter(sleep_total_24<=0.3) %>% 
  arrange(desc(sleep_total_24))
msleep24
```

Did `dplyr` do what we expected? How do we check our output? Remember, just because your code runs it doesn't mean that it did what you intended.
```{r}
summary(msleep24)
```

Try out the new function `skim()` as part of the `skimr` package.
```{r}
skim(msleep24)
```

Histograms are also a quick way to check the output.
```{r}
hist(msleep24$sleep_total_24)
```

## `summarize()`
`summarize()` will produce summary statistics for a given variable in a data frame. For example, if you are asked to calculate the mean of `sleep_total` for large and small mammals you could do this using a combination of commands, but it isn't very efficient or clean. We can do better!  
```{r}
head(msleep)
```

For example, if we define "large" as having a `bodywt` greater than 200 then we get the following:
```{r}
large <- msleep %>% 
  select(name, genus, bodywt, sleep_total) %>% 
  filter(bodywt > 200) %>% 
  arrange(desc(bodywt))
large
```

```{r}
mean(large$sleep_total)
```

We can accomplish the same task using the `summarize()` function to make things cleaner.
```{r}
msleep %>% 
  filter(bodywt>200) %>% 
  summarise(mean_sleep_large=mean(sleep_total))
```
```{r}
msleep %>% 
  filter(bodywt<10) %>% 
  summarise(mean_sleep_small=mean(sleep_total)) #mean of small animal sleep because you already picked out the weight class
```

You can also combine functions to make useful summaries for multiple variables.
```{r}
msleep %>% 
  filter(bodywt>200) %>% 
  summarise(mean_sleep_large=mean(sleep_total),
            min_sleep_large=min(sleep_total),
            max_sleep_large=max(sleep_total),
            mean_sleep_large=mean(sleep_total),
            sd_sleep_large=sd(sleep_total),
            total=n())
```

## Practice
1. What is the mean, min, and max `bodywt` for the taxonomic order Primates? Provide the total number of observations.
```{r}
names(msleep)
```


```{r}
msleep %>% 
  filter(order=="Primates") %>% 
  summarise(mean_body_pri=mean(bodywt),
            min_body_pri=min(bodywt),
            max_body_pri=max(bodywt),
            total=n())
```

`n_distinct()` is a very handy way of cleanly presenting the number of distinct observations. Here we show the number of distinct genera over 100 in body weight.  
#NEED
Notice that there are multiple genera with over 100 in body weight.
```{r}
msleep %>% 
  filter(bodywt > 100)
```

n_distinct() is a very handy way of cleanly presenting the number of distinct observations. Here we show the number of distinct genera over 100 in body weight.
```{r}
msleep %>% 
  summarise(n_genera=n_distinct(genus)) #this is going to count the number of genera in msleep
```
```{r}
glimpse(msleep) #see there is 83 rows but 77 genera. There are 5 duplicate so we filtered that out
```

There are many other useful summary statistics, depending on your needs: sd(), min(), max(), median(), sum(), n() (returns the length of a column), first() (returns first value in a column), last() (returns last value in a column) and n_distinct() (number of distinct values in a column).

## Practice
1. How many genera are represented in the msleep data frame?
```{r}

```

```{r}

```

2. What are the min, max, and mean `sleep_total` for all of the mammals? Be sure to include the total n.
```{r}
msleep %>% 
  summarise(mean_bodywt=mean(bodywt),
            min_bodywt=min(bodywt),
            max_bodywt=max(bodywt),
            total=n())
```

## `group_by()`##### GROUP_BY
The `summarize()` function is most useful when used in conjunction with `group_by()`. Although producing a summary of body weight for all of the mammals in the data set is helpful, what if we were interested in body weight by feeding ecology?
```{r}
msleep %>%
  group_by(vore) %>% #we are grouping by feeding ecology, a categorical variable
  summarize(min_bodywt = min(bodywt),
            max_bodywt = max(bodywt),
            mean_bodywt = mean(bodywt),
            total=n())
```

## Practice
1. Calculate mean brain weight by taxonomic order in the msleep data.
```{r}
msleep %>% 
  group_by(order) %>% 
  summarise(mean_brain=mean(brainwt),
            total=n())
```

2. What does `NA` mean? How are NA's being treated by the summarize function?
```{r}
#carnivora has some alues for brain weight but has some NA. but it only spits out na as a value. 1 NA can screw the whole thing up

```

3. Try running the code again, but this time add `na.rm=TRUE`. What is the problem with Cetacea? Compare this to Carnivora. 
```{r}
msleep %>% 
  group_by(order) %>% 
  summarise(mean_brain=mean(brainwt, na.rm=T),
            total=n())
```
NaN because they is no data

```{r}
msleep %>% 
  filter(order=="Cetacea") %>% 
  select(order,genus,brainwt)
```

## That's it! Let's take a break and then move on to part 2! 

-->[Home](https://jmledford3115.github.io/datascibiol/)  